---
title: "R Basics"
output:
  revealjs::revealjs_presentation:
    theme: white
    center: true
    transition: none
    incremental: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```
# R Intro

## Reading Data files
The ability to datasets into R is an essential skill. For this class, most of the files will be on the course webpage and can be directly downloaded using `read_csv`. Consider the Seattle Housing dataset from the previous lecture.

```{r read.data}
Seattle <- read_csv('http://math.montana.edu/ahoegh/teaching/stat408/datasets/SeattleHousing.csv')
```

## Reading Data files
We can also use the `read.csv()` function.

```{r}
Seattle2 <- read.csv('http://math.montana.edu/ahoegh/teaching/stat408/datasets/SeattleHousing.csv')
```

## Viewing Data files
A common function that we will use is `head`, which shows the first few rows of a data frame.

```{r}
head(Seattle)
```

## Viewing Data files
Another option is `glimpse`, which details the columns of a data frame (tibble).

```{r}
glimpse(Seattle)
```


## Viewing Data files

```{r}
head(Seattle2)
```

## Viewing Data files

```{r}
glimpse(Seattle2)
```


## Activity: Reading excel files

The `readxl` package makes importing excel data files easy. 

- Find an excel dataset (https://catalog.data.gov/dataset/?q=&_res_format_limit=0&res_format=EXCEL)
- Download the dataset
- Read this into R, pay attention to the sheet names.
- Extra. Write the Seattle dataset from R to CSV and then XLS.



# R Data Structures
## Data structure Overview
R has four common types of data structures:

- Vectors
- Matrices (and Arrays) 
- Lists
- Data Frames (including tibbles)

## Data structure Overview
The base data structures in R can be organized by dimensionality and whether they are homogenous.

Dimension     | Homogenous     | Heterogenous
------------- | -------------  | -------------
1d            | Vector         | List
2d            | Matrix         | Data Frame
no d          | Array          | 


## Vector Types
There are four common types of vectors: logical, integer, double (or numeric), and character. The `c()` function is used for combining elements into a vector

```{r vectors}
dbl <- c(1,2.5,pi)
int <- c(1L,4L,10L)
log <- c(TRUE,FALSE,F,T)
char <- c('this is','a character string')
```



## Vector Types
They type of vector can be identified using the `typeof()` function. Note that only a single data type is allowed.
```{r vector.type}
  typeof(dbl)
  comb <- c(char,dbl)
  typeof(comb)
  comb
```


## Exercise: Vectors

Create a vector with your first, middle, and last names.


## Solution: Vectors

Create a vector with your first, middle, and last names.

```{r}
andy.names <- c("Andrew","Blake","Hoegh")
andy.names
```

## Logical Values in R (Conditions)
Logical values are `TRUE` or `FALSE`. Evaluate these expressions:
```{r}
TRUE
pi > 3
pi == 3.14
```


## Exercise: Conditions in R
Note that `&` is an and operator

```{r, eval=FALSE}
pi > 3 & pi < 3.5
c(1,3,5,7) %in% 1:3
1:3 %in% c(1,3,5,7)
```

## Solutions: Conditions in R: Evaluated
\footnotesize
```{r}
pi > 3 & pi < 3.5
c(1,3,5,7) %in% 1:3
1:3 %in% c(1,3,5,7)
```


## Data Frame Overview
A data frame:

>- is the most common way of storing data in R
>- is like a matrix with rows-and-column structure; however, unlike a matrix each column may have a different mode
>- in a technical sense, a data frame is a list of equal-length vectors.

```{r df}
df <- data.frame(x = 1:3, y = c('a','b','c'))
kable(df)
```

## tibble()
A modern data frame can be constructed using the `tibble()` command.

>- The `read_csv` command creates a tibble rather than a data.frame.

>- The tibble includes the type of each vector, and only prints a certain number of rows/columns.

```{r}
tibble1 <- tibble(x = 1:3, y = c('a','b','c')); tibble1
```


# Subsetting
 
## Vector Subsetting: I
Subsetting allows you to extract elements from an object.

```{r }
num.vec <- seq(from = 1, to = 9, by = 1); num.vec
num.vec[1:3]
num.vec[c(1,5,8)]
```

## Vector Subsetting: II
Subsetting also works with negative values or expressions.
```{r subset2}
num.vec[-5]
num.vec[num.vec != 6]
num.vec[num.vec > 5]
```

## Vector Subsetting: III
Another possibility is to use logical values directly.
```{r subset3}
num.vec > 5
num.vec[num.vec > 5]
num.vec[rep(c(TRUE,FALSE,TRUE),each=3)]
```


## Data Frame Subsetting: I
The same ideas apply to data frames, but the indices now constitute rows and columns of the data frame.
```{r subset.df}
df <- data.frame(x=1:3, y=3:1, z=c('a','b','c'))
df[,1]
df[-1,c(2:3)]
```


## Data Frame Subsetting: II
The `filter()` and `select()` functions in the `dplyr` package (in the tidyverse) can also be used for subsetting.
```{r subset.df3}
df |>
  select(x)
filter(df, x > 1)
```

## Exercise: Subsetting
1. Create a new data frame that only includes houses worth more than $1,000,000.

2. From this new data frame what is the average living square footage of houses. 

## Exercise: Subsetting - Solutions1

Create a new data frame that only includes houses worth more than $1,000,000.

```{r}
expensive_houses <- Seattle |>
  filter(price > 1000000)
```

## Exercise: Subsetting - Solutions2

From this new data frame what is the average living square footage of houses. Hint columns in a data.frame can be indexed by `Seattle$sqft_living`

```{r}
mean(expensive_houses$sqft_living)
Seattle |>
  filter(price > 1000000) |>
  summarize(ave_size = mean(sqft_living))
```


## Lists
Consider the two lists 

```{r}
msu.info <- list( name = c('Waded Cruzado','Andy Hoegh'), 
         degree.from = c('University of Texas at Arlington','Virginia Tech'),
         job.title = c('President', 'Assistant Professor of Statistics'))

msu.info2 <- list(c('Waded Cruzado','University of Texas at Arlington',
                     'President'), c('Andy Hoegh',
                  'Virginia Tech','Assistant Professor of Statistics'))
```

## List Output

```{r, mysize=TRUE, size='\\tiny'}

msu.info

msu.info2
```

## Lists - indexing
With the current lists we can index elements using the double bracket `[[ ]]` notation or if names have been initialized, those can be used too.

So the first element of each list can be indexed
```{r, mysize=TRUE, size='\\tiny'}
msu.info[[1]]
msu.info$name
```

## Exercise: Lists 
Explore the indexing with these commands.
```{r, mysize=TRUE, size='\\tiny', eval=F}
msu.info <- list( name = c('Waded Cruzado','Andy Hoegh'), 
         degree.from = c('University of Texas at Arlington','Virginia Tech'),
         job.title = c('President', 'Associate Professor of Statistics'))
msu.info[1]
msu.info[[1]]
msu.info$name[2]
msu.info[1:2]
unlist(msu.info)
```

## Solution: Lists 1
```{r, mysize=TRUE, size='\\tiny'}
msu.info[1]
msu.info[[1]]
msu.info$name[2]
```

## Solution: Lists 2
```{r, mysize=TRUE, size='\\tiny'}
msu.info[1:2]
unlist(msu.info)
```


## Lists  - nested lists

```{r, mysize=TRUE, size='\\footnotesize'}

list(list('a','b'),list('c','d'))
```


## Arrays

Arrays are a general form a matrix, but have a higher dimension.

```{r, mysize=TRUE, size='\\tiny'}
array.1 <- array(1:8, dim=c(2,2,2)); array.1
array.1[2,2,1]
```

## Exercise: Arrays

Create an array of dimension 2 x 2 x 3, where each of the three 2 x 2 subarray (or matrix) is the Identity matrix.

## Solution: Arrays
Create an array of dimension 2 x 2 x 3, where each of the three 2 x 2 subarray (or matrix) is the Identity matrix.

```{r, mysize=TRUE, size='\\tiny'}
array(c(1,0,0,1), dim = c(2,2,3))
```

